### 2021-03-19 ch/2-related-work.tex
Nearly there

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2012fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalization that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

An instance of this setting has been formalised by \citet{blum2020recovering}.
In their formalization, the training set starts out unbiased,
but then, for samples with \(s=0\), class labels \(y\) are flipped from 1 to 0 with probability \(\nu\).
In addition, samples with \(s=0\) and \(y=0\) are dropped with probability \(1-\beta_\mathit{NEG}\)
and those with \(s=0\) and \(y=1\) are dropped with probability \(1-\beta_\mathit{POS}\).
Meanwhile, the test remains unbiased and is the basis for evaluating our models.

@@

\section{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
Their first fundamental conclusion
is that it is not sufficient to remove the protected attribute
(which should not be used as a basis for prediction) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

\section{Modification of labels for Fairness enforcement}\label{modification-of-labels-for-fairness-enforcement}
Independently of \citet{pedreshi2008discrimination}, but with a similar intention,
\citet{kamiran2009classifying} considered discrimination in classification.
Their goal is to modify the dataset so that any classifier trained on it will be ``fair'',
which is measured with a fairness metric on the test set predictions.
The test set is drawn from the same distribution as the training set.
To demonstrate their method, they use the ``German credit dataset''~\citep{Dua:2017}
where the task is to classify people as a good or a bad credit risk.

@@

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before.
They present a different technique based on giving training examples different weights
instead of re-labelling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex}.
In addition to Na\"ive Bayes,
two classifiers based on nearest neighbours and a decision tree classifier are used.
The comparison shows that changing the labels leads to lower dependency (discrimination)
than reweighting the examples.
However, reweighting the examples leads to lower dependency than the baseline.
% The difference in accuracy is minor.

\section{Fair classifiers}\label{fair-classifiers}
Another common approach is to put fairness \emph{constraints} on the classifier
instead of manipulating the dataset.
\citet{calders2010three} proposes three such approaches,
which potentially give more control over the prediction bias than changing the datasets.
All three approaches are based on Na\"ive Bayes
and again aim to enforce \ac{DP} in the predictions as defined before.
The first approach shifts the predicted probabilities in a post-processing step
after training the classifier normally.
To this end, the sensitive attribute is treated differently
than the other features in the Na\"ive Bayes model.
Instead of the class being the cause of the sensitive attribute,
the sensitive attribute is regarded as one cause for the class.
This is a significant change in perspective that enables a better intuition about the problem.
In this modified Na\"ive Bayes,
the conditional probability \(P(y|s)\) is then modified in the already trained model
until the discrimination score is below a given threshold.
Care is taken to not distort the predicted distribution too much with respect to the unfair model.
The end result is a model that is (ideally) unbiased
but still relies on the sensitive attribute for predictions.
This is unworkable if the sensitive attributes are not always available when making predictions.
Nevertheless, the algorithm can be considered an improvement over manipulating the dataset,
as the resulting bias can be controlled to a much finer degree.

@@

The third methods introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximization
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
% This method seems to have been the most sophisticated at that point in time.

\citet{calders2010three} present experimental results for the three methods on an artificial dataset
and the Adult/Census Income data that \citet{calders2009building} used before.
The artificial data has biased and also unbiased class labels
which are generally not available for real-world datasets
and conforms to the assumptions made for the third model.
% Artificial datasets generally have the advantage that the bias can be controlled very precisely,
% but there is always the danger of tailoring the dataset too closely to the proposed algorithm.
To the authors surprise, on both datasets the two simpler methods outperformed the third method
that was based on fair latent class labels;
the two first methods perform approximately equally well.
% The authors mention the possibility of a better fairness measure.

@@

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2012fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2012fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimised together with the other parts of the objective function.
% This strategy leaves most of the work to the optimization that would need to be run anyway.
A factor in front of the regularization term determines how much to value fairness over accuracy.

% In a comparison between \citet{kamishima2012fairness}'s algorithm
% and the Na\"ive Bayes method with two different models for the sensitive attributes from \citet{calders2010three},
% the latter performed better on the Adult/Census Income dataset
% (higher accuracy and lower discrimination score).
% However, the former has the advantage
% that the sensitive attribute does not necessarily have to be known for making predictions
% and the trade-off between accuracy and fairness can be easily controlled
% with the factor for the regularization term.

@@

The proposed method is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
% This is similar to how \citet{kamishima2012fairness} used the fairness criterion as a regulariser.
% However, \citet{dwork2012fairness} intend to realise it as a strict constraint.
% Furthermore, they show that their criterion is compatible with Demographic Parity
% but does not guarantee it in general.
% When the two demographic groups are similar according to the individual similarity measure,
% then the Lipschitz condition will lead to a good approximation of Demographic Parity.
% If the demographic groups are very different,
% then the Lipschitz criterion is essentially incompatible with Demographic Parity.

@@

\section{Fair representations}\label{sec:fair-representation}
\citet{zemel2013learning} approach the problem of biased training data more directly
by seeking to transform the features such that all traces of the sensitive attribute are removed,
but so that they still carry the information necessary for predicting the class label.
% \citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
% That is, the features of the data points are transformed such that all bias is removed
This is different from \citet{kamiran2009classifying} where the dataset was transformed as well,
but there, the change was to the \emph{labels}.
The idea of transforming the features assumes that the main problem with the data
are unwanted dependencies between $s$ and the other features (called \emph{latent prejudice} in \citet{kamishima2012fairness}),
and that the training labels are mostly unproblematic.
The intended result is then that classifiers trained on the transformed data will make $s$-invariant predictions `by default',
because they do not know about $s$.

@@

In order to create a fair dataset,
the authors present an algorithm that considers every feature individually
and shifts the values such that the distributions \(P(z=z'|s=0)\) and \(P(z=z'|s=1)\) are identical
(\(z\) refers to the shifted values, \(x\) refers to the original values).
This shift retains the ordering of the data points with regard to that feature.
This is to ensure that \(z\) can still be used to predict \(y\).
The method only works on numerical features.
The paper contains two other algorithms for removing unfairness which are not as invasive,
meaning some amount of unfairness remains, but the ability to predict is improved.
They are referred to as \emph{Partial Repair} algorithms, as opposed to full repair.
This is the fairness-accuracy trade-off that basically all works in this area consider.
Both of the other methods try to preserve the ranking of the data points
with respect to the individual features
while minimizing the distance of the distributions for the different groups.

@@

While the algorithms by \citet{feldman2015certifying} are manually constructed and explicit,
\citet{louizos2016variational} use an approach
that falls more in the area of end-to-end learning where fewer hand-crafted algorithms are used.
The method is based on deep variational autoencoders (VAE),
with an encoder and a decoder that are both modelled as deep neural networks.
The encoder produces the distribution of the latent (fair) representation \(z\)
from the original features \(x\) and the sensitive attribute \(s\).
The decoder recovers the distribution of \(x\) from \(z\) and \(s\).
By choosing a factorised prior \(P(s)P(z)\), a separation between \(s\) and \(z\) is encouraged.

@@

\section{Other fairness criteria}%
\label{improved-definitions-of-fairness}
The early literature on fairness in machine learning used only \ac{DP}
and si\-mi\-la\-ri\-ty-based fairness to measure the bias in predictions.
\citet{kleinberg2016inherent} formalised three different group fairness conditions.
The authors consider a scenario where the data points \(x\) are sorted into bins \(b\)
and each bin is associated with a prediction score \(f = P(\hat{y}=1|b)\)
where \(\hat{y}\) is the predicted class label and \(x\) the (non-sensitive) features.
\begin{enumerate}
\item
  Calibration within groups:
  If the prediction score for a given bin is \(f\),
  then when considering all the data points with group \(s\) in the bin,
  a fraction of \(f\) of those should have the class label \(y=1\).
  In other words, the prediction score \(f\) is well calibrated with respect to group \(s\).
  This should be the case for all groups.
\item
  Balance for the negative class:
  The true negative rate should be the same for both groups: \(P(\hat{y}=0|y=0,s=0) = P(\hat{y}=0|y=0,s=1)\).
\item
  Balance for the positive class:
  The true positive rate should be the same for both groups: \(P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)\).
\end{enumerate}
% These are formalizations of things that have been proposed as definitions of fairness.
The first criterion essentially ensures that the predictor works correctly for both groups.
This prevents a classifier from predicting one group correctly
but always returning a negative answer for the other group.
The second and third put emphasis on negative and positive class labels respectively.
In criterion 2, we allow the classifier to mis-classify those data points with \(y=0\),
but we want to make sure that the misclassification rate is the same for both groups.
In other words, members of the different groups have the same chance
to get a correct classification if they should receive a negative classification.
The same holds for criterion 3 and positive classifications (\(y=1\)).

@@

In order to construct a fair classifier out of a \emph{binary predictor}
(giving only binary predictions) via post-processing,
the output is randomised in such a way as to remove the bias.
If the overall probability for a positive predictions of the unfair classifier is given by \(P(\hat{y}=1|s)\),
then the randomised probability for the fair positive predictions \(\tilde{y} = 1\) is given by:
\begin{align}
  \label{eq:hardt}
  P(\tilde{y}=1| x, s) = \sum\limits_{y' \in \{0, 1\}} P(\tilde{y} = 1| \hat{y}=y', s) 
  P(\hat{y}=y'| x, s)
\end{align}
There are 4 free parameters \(P(\tilde{y} = 1| \hat{y}=y', s=s')\) with \(y' \in \{0, 1\}\) and \(s' \in \{0, 1\}\).
As an example, if the unfair predictor predicts \(\hat{y} =0\) for an input with \(s=0\),
then the fair predictor predicts \(\tilde{y} =1\) with probability \(P(\tilde{y} = 1| \hat{y}=0, s=0)\)
which might be non-zero.
In addition to the fairness condition, we also want to enforce accuracy.
To that end, we introduce a loss function \(\ell (\tilde{y}, y)\)
that quantifies the cost of predicting the wrong class label.
The final optimization problem for the 4 free parameters is then to minimise \(\ell\)
under the constraint of \ac{EOpp} or \ac{EOdds}
and the constraint that the parameters must be valid probabilities.

For a predictor that outputs a \emph{score function},
the post-processing step consists of choosing differing thresholds for \(s=0\) and \(s=1\),
such that the predictions become fair.
If \(f\) is the score, then for a given threshold \(t\) we predict \(\hat{y} = 1\) if \(f > t\).
The two thresholds (one for each group) are found by minimizing \(\ell\) with the fairness constraints, as before.
To satisfy the constraint, it might be necessary to randomise the result.
This is done by using two constraints per group:
if \(f\) is above both, the result is \(\hat{y} = 1\), if it is below both, \(\hat{y} =0\)
and if \(f\) is between the thresholds, then the result is chosen at random.
This method -- as well as the one before --
requires the sensitive attribute for all predictions as it has to choose the right threshold.

@@

\section{Refinement of previous ideas}%
\label{sec:refinement-on-previous-ideas-for-fair-classification-and-fair-representations}
\subsection{Fair classifiers continued}
Following the initial publications,
several works refined the ideas that had been presented there.
In a very influential work by \citet{zafar2017fairnessconstraints}
present a more efficient way to train a fair classifier.
The authors explicitly discuss the tension
between the two goals of making fair predictions and not making use of the sensitive attributes.
Using sensitive attributes to make predictions
is referred to as \emph{Disparate Treatment} by the authors.
Algorithms like the one by \citet{calders2009building} use the sensitive
attribute during prediction to achieve a very high degree of fairness, but this can have a
lot of problems from a legal perspective and can also easily go wrong.
(Making use of \(s\) comes close to \emph{direct discrimination} discussed above.)
Thus, \citet{zafar2017fairnessconstraints} set themselves the goal of achieving fairness while making use of the sensitive attributes
as little as possible, i.e.~avoiding Disparate Treatment.

@@

% Combining privileged learning and the \ac{MMD} fairness criterion,
% the task becomes a multi-objective optimization for an SVM.
% The MMD criterion is generally non-convex but there are several ways to deal with this.
% The authors find the best algorithm (low discrimination, high accuracy)
% with an evolutionary multi-objective optimization.
% The evolutionary algorithm produces a number of variants of the classifier
% with different trade-offs between classification error and discrimination.
% As in \citet{zafar2017fairnesstreatment} before,
% the Adult/Census Income dataset is used to test Demographic Parity
% and the ProPublica/COMPAS dataset is used for Equality of Opportunity and Equalised Odds.
% \citet{zafar2017fairnesstreatment} was used as a baseline
% but the implementation failed on the Adult/Census Income dataset.
% On the ProPublica/COMPAS dataset,
% the proposed algorithm performed within error just as well as the fair baseline algorithm.

@@

% \citet{kilbertus2018blind} recently proposed a mechanism
% where the sensitive attributes are encrypted to avoid being used for discrimination.
% \citet{bechavod2017penalizing} use (as some others before) a regularization approach to train fair classifiers.

>>>>

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2012fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalisation that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

An instance of this setting has been formalised by \citet{blum2020recovering}.
In their formalisation, the training set starts out unbiased,
but then, for samples with \(s=0\), class labels \(y\) are flipped from 1 to 0 with probability \(\nu\).
In addition, samples with \(s=0\) and \(y=0\) are dropped with probability \(1-\beta_\mathit{NEG}\)
and those with \(s=0\) and \(y=1\) are dropped with probability \(1-\beta_\mathit{POS}\).
Meanwhile, the test remains unbiased and is the basis for evaluating our models.

@@

\section{Foundations of algorithmic fairness}\label{sec:foundations}
This section discusses the publications that lay the ground work for the algorithmic fairness literature.
More recent publications are discussed in \secref{sec:refinement-on-previous-ideas-for-fair-classification-and-fair-representations}.

\subsection{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
Their first fundamental conclusion
is that it is not sufficient to remove the protected attribute
(which should not be used as a basis for prediction) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

\subsection{Modification of labels for fairness enforcement}%
\label{modification-of-labels-for-fairness-enforcement}
Independently of \citet{pedreshi2008discrimination}, but with a similar intention,
\citet{kamiran2009classifying} considered discrimination in classification.
Their goal is to modify the dataset so that any classifier trained on it will be ``fair'',
which is measured with a fairness metric on the test set predictions.
The test set is drawn from the same distribution as the training set.
To demonstrate their method, they use the ``German credit dataset''~\citep{Dua:2017}
where the task is to classify people as a good or a bad credit risk.

@@

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before.
They present a different technique based on giving training examples different weights
instead of re-labelling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex} (``gender'' in some publications).
% In addition to Na\"ive Bayes,
% two classifiers based on nearest neighbours and a decision tree classifier are used.
% The comparison shows that changing the labels leads to lower dependency (discrimination)
% than reweighting the examples.
% However, reweighting the examples leads to lower dependency than the baseline.
% The difference in accuracy is minor.

\subsection{Fair classifiers}\label{fair-classifiers}
Another common approach is to put fairness \emph{constraints} on the classifier
instead of manipulating the dataset.
\citet{calders2010three} proposes three such approaches,
which potentially give more control over the prediction bias than changing the datasets.
All three approaches are based on Na\"ive Bayes
and again aim to enforce \ac{DP} in the predictions as defined before.
The first approach shifts the predicted probabilities in a post-processing step
after training the classifier normally.
To this end, the sensitive attribute is treated differently
than the other features in the Na\"ive Bayes model.
Instead of the class being the cause of the sensitive attribute,
the sensitive attribute is regarded as one cause for the class.
This is a significant change in perspective that enables a better intuition about the problem.
In this modified Na\"ive Bayes,
the conditional probability \(P(y|s)\) is then modified in the already trained model
until the discrimination score is below a given threshold.
Care is taken to not distort the predicted distribution too much with respect to the unfair model.
The end result is a model that is (ideally) unbiased
but still relies on the sensitive attribute for predictions.
This is unworkable if the sensitive attributes are not always available when making predictions.
Nevertheless, the algorithm can be considered an improvement over manipulating the dataset,
as the resulting bias can be controlled to a much finer degree.

@@

The third methods introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximisation
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
% This method seems to have been the most sophisticated at that point in time.

\citet{calders2010three} present experimental results for the three methods on an artificial dataset
and the Adult/Census Income data that \citet{calders2009building} used before.
The artificial data has biased and also unbiased class labels
which are generally not available for real-world datasets
and conforms to the assumptions made for the third model.
% Artificial datasets generally have the advantage that the bias can be controlled very precisely,
% but there is always the danger of tailoring the dataset too closely to the proposed algorithm.
% To the authors surprise,
On both datasets, the two simpler methods outperform the third method
that is based on fair latent class labels;
the two first methods perform approximately equally well.
% The authors mention the possibility of a better fairness measure.

@@

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2012fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2012fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimised together with the other parts of the objective function.
% This strategy leaves most of the work to the optimisation that would need to be run anyway.
A factor in front of the regularisation term determines how much to value fairness over accuracy.

% In a comparison between \citet{kamishima2012fairness}'s algorithm
% and the Na\"ive Bayes method with two different models for the sensitive attributes from \citet{calders2010three},
% the latter performed better on the Adult/Census Income dataset
% (higher accuracy and lower discrimination score).
% However, the former has the advantage
% that the sensitive attribute does not necessarily have to be known for making predictions
% and the trade-off between accuracy and fairness can be easily controlled
% with the factor for the regularisation term.

@@

The proposed method is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimisation.
% This is similar to how \citet{kamishima2012fairness} used the fairness criterion as a regulariser.
% However, \citet{dwork2012fairness} intend to realise it as a strict constraint.
% Furthermore, they show that their criterion is compatible with Demographic Parity
% but does not guarantee it in general.
% When the two demographic groups are similar according to the individual similarity measure,
% then the Lipschitz condition will lead to a good approximation of Demographic Parity.
% If the demographic groups are very different,
% then the Lipschitz criterion is essentially incompatible with Demographic Parity.

@@

\subsection{Fair representations}\label{sec:fair-representation}
\citet{zemel2013learning} approach the problem of biased training data more directly
by seeking to transform the features such that all traces of the sensitive attribute are removed,
but so that they still carry the information necessary for predicting the class label.
% \citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
% That is, the features of the data points are transformed such that all bias is removed
This is different from \citet{kamiran2009classifying} where the dataset was transformed as well,
but there, the change was to the \emph{labels}.
The idea of transforming the features assumes that the main problem with the data
are unwanted dependencies between $s$ and the other features (called \emph{latent prejudice} in \citet{kamishima2012fairness}),
and that the training labels are mostly unproblematic.
The intended result is then that classifiers trained on the transformed data will make $s$-invariant predictions `by default',
because they do not know about $s$.

@@

In order to create a fair dataset,
the authors present an algorithm that considers every feature individually
and shifts the values such that the distributions \(P(z=z'|s=0)\) and \(P(z=z'|s=1)\) are identical
(\(z\) refers to the shifted values, \(x\) refers to the original values).
This shift retains the ordering of the data points with regard to that feature.
This is to ensure that \(z\) can still be used to predict \(y\).
The method only works on numerical features.
The paper contains two other algorithms for removing unfairness which are not as invasive,
meaning some amount of unfairness remains, but the ability to predict is improved.
They are referred to as \emph{Partial Repair} algorithms, as opposed to full repair.
This is the fairness-accuracy trade-off that basically all works in this area consider.
Both of the other methods try to preserve the ranking of the data points
with respect to the individual features
while minimising the distance of the distributions for the different groups.

@@

While the algorithms by \citet{feldman2015certifying} are manually constructed and explicit,
\citet{louizos2016variational} use an approach
that falls more in the area of end-to-end learning where fewer hand-crafted algorithms are used.
The method is based on deep \acp{VAE},
with an encoder and a decoder that are both modelled as deep neural networks.
The encoder produces the distribution of the latent (fair) representation \(z\)
from the original features \(x\) and the sensitive attribute \(s\).
The decoder recovers the distribution of \(x\) from \(z\) and \(s\).
By choosing a factorised prior \(P(s)P(z)\), a separation between \(s\) and \(z\) is encouraged.

@@

\subsection{Other fairness criteria}%
\label{improved-definitions-of-fairness}
The early literature on fairness in machine learning used only \ac{DP}
and si\-mi\-la\-ri\-ty-based fairness to measure the bias in predictions.
\citet{kleinberg2016inherent} formalised three different group fairness conditions.
The authors consider a scenario where the data points \(x\) are sorted into bins \(b\)
and each bin is associated with a prediction score \(f = P(\hat{y}=1|b)\)
where \(\hat{y}\) is the predicted class label and \(x\) the (non-sensitive) features.
\begin{enumerate}
\item
  Calibration within groups:
  If the prediction score for a given bin is \(f\),
  then when considering all the data points with group \(s\) in the bin,
  a fraction of \(f\) of those should have the class label \(y=1\).
  In other words, the prediction score \(f\) is well calibrated with respect to group \(s\).
  This should be the case for all groups.
\item
  Balance for the negative class:
  The true negative rate should be the same for both groups: \(P(\hat{y}=0|y=0,s=0) = P(\hat{y}=0|y=0,s=1)\).
\item
  Balance for the positive class:
  The true positive rate should be the same for both groups: \(P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)\).
\end{enumerate}
% These are formalisations of things that have been proposed as definitions of fairness.
The first criterion essentially ensures that the predictor works correctly for both groups.
This prevents a classifier from predicting one group correctly
but always returning a negative answer for the other group.
The second and third put emphasis on negative and positive class labels respectively.
In criterion 2, we allow the classifier to mis-classify those data points with \(y=0\),
but we want to make sure that the misclassification rate is the same for both groups.
In other words, members of the different groups have the same chance
to get a correct classification if they should receive a negative classification.
The same holds for criterion 3 and positive classifications (\(y=1\)).

@@

In order to construct a fair classifier out of a \emph{binary predictor}
(giving only binary predictions) via post-processing,
the output is randomised in such a way as to remove the bias.
If the overall probability for a positive predictions of the unfair classifier is given by \(P(\hat{y}=1|s)\),
then the randomised probability for the fair positive predictions \(\tilde{y} = 1\) is given by:
\begin{align}
  \label{eq:hardt}
  P(\tilde{y}=1| x, s) = \sum\limits_{y' \in \{0, 1\}} P(\tilde{y} = 1| \hat{y}=y', s) 
  P(\hat{y}=y'| x, s)
\end{align}
There are 4 free parameters \(P(\tilde{y} = 1| \hat{y}=y', s=s')\) with \(y' \in \{0, 1\}\) and \(s' \in \{0, 1\}\).
As an example, if the unfair predictor predicts \(\hat{y} =0\) for an input with \(s=0\),
then the fair predictor predicts \(\tilde{y} =1\) with probability \(P(\tilde{y} = 1| \hat{y}=0, s=0)\)
which might be non-zero.
In addition to the fairness condition, we also want to enforce accuracy.
To that end, we introduce a loss function \(\ell (\tilde{y}, y)\)
that quantifies the cost of predicting the wrong class label.
The final optimisation problem for the 4 free parameters is then to minimise \(\ell\)
under the constraint of \ac{EOpp} or \ac{EOdds}
and the constraint that the parameters must be valid probabilities.

For a predictor that outputs a \emph{score function},
the post-processing step consists of choosing differing thresholds for \(s=0\) and \(s=1\),
such that the predictions become fair.
If \(f\) is the score, then for a given threshold \(t\) we predict \(\hat{y} = 1\) if \(f > t\).
The two thresholds (one for each group) are found by minimising \(\ell\) with the fairness constraints, as before.
To satisfy the constraint, it might be necessary to randomise the result.
This is done by using two constraints per group:
if \(f\) is above both, the result is \(\hat{y} = 1\), if it is below both, \(\hat{y} =0\)
and if \(f\) is between the thresholds, then the result is chosen at random.
This method -- as well as the one before --
requires the sensitive attribute for all predictions as it has to choose the right threshold.

@@

\section{Recent developments in algorithmic fairness}%
\label{sec:refinement-on-previous-ideas-for-fair-classification-and-fair-representations}
This section discusses more recent publications from the fairness literature.

\subsection{Fair classifiers continued}
Following the initial publications,
several works refined the ideas that had been presented there.
A very influential work by \citet{zafar2017fairnessconstraints}
presents a more efficient way to train a fair classifier.
The authors explicitly discuss the tension
between the two goals of making fair predictions and not making use of the sensitive attributes.
Using sensitive attributes to make predictions
is referred to as \emph{Disparate Treatment} by the authors.
Algorithms like the one by \citet{calders2009building} use the sensitive
attribute during prediction to achieve a very high degree of fairness, but this can have a
lot of problems from a legal perspective and can also easily go wrong.
(Making use of \(s\) comes close to \emph{direct discrimination} discussed above.)
Thus, \citet{zafar2017fairnessconstraints} set themselves the goal of achieving fairness while making use of the sensitive attributes
as little as possible, i.e.~avoiding Disparate Treatment.

@@

% Combining privileged learning and the \ac{MMD} fairness criterion,
% the task becomes a multi-objective optimisation for an SVM.
% The MMD criterion is generally non-convex but there are several ways to deal with this.
% The authors find the best algorithm (low discrimination, high accuracy)
% with an evolutionary multi-objective optimisation.
% The evolutionary algorithm produces a number of variants of the classifier
% with different trade-offs between classification error and discrimination.
% As in \citet{zafar2017fairnesstreatment} before,
% the Adult/Census Income dataset is used to test Demographic Parity
% and the ProPublica/COMPAS dataset is used for Equality of Opportunity and Equalised Odds.
% \citet{zafar2017fairnesstreatment} was used as a baseline
% but the implementation failed on the Adult/Census Income dataset.
% On the ProPublica/COMPAS dataset,
% the proposed algorithm performed within error just as well as the fair baseline algorithm.

@@

% \citet{kilbertus2018blind} recently proposed a mechanism
% where the sensitive attributes are encrypted to avoid being used for discrimination.
% \citet{bechavod2017penalizing} use (as some others before) a regularisation approach to train fair classifiers.
